# Introduction to Probability {#probability}

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```



```{r}
#note: some of these aren't used for the code that I run, but for other parts I have hashed out.
library(tidyverse)
library(combinat)
library(gtools)
library(readxl)
library(ggplot2)
library(ggpubr)
library(mosaic)

library(dplyr)
library(gganimate)
library(gifski)
library(png)
library(installr)
```

In this lesson, we will learn about the classical theory of probability, through a set of worked examples in R.

## Simulating Coin Flips

The first thing we'll do is simulate some fair coin flips (we'll come back to this later as well)

Note: This is virtually copied verbatim from: <https://rpubs.com/pgrosse/545948>

Remember, this is another random process, so results will differ each time slightly

```{r}
#simulate 5000 flips of a fair coin
#create data frame for trial number and outcome of an individual coin flip
flips <- sample(c(0, 1), 5000, replace = TRUE)
flips <- matrix(flips, ncol = 1)
flips <- as.data.frame(flips)
Trial <- seq(1, 5000, 1)
Trial <- as.data.frame(Trial)
flipsim <- cbind(flips, Trial)
colnames(flipsim) <- c("Heads", "Trial")
```

```{r}
#calculate cumulative heads at the end of each trial
flipsim[,"Cum_Heads"] <- cumsum(flipsim$Heads)
flipsim <- flipsim %>% mutate(Pct_Heads = Cum_Heads/Trial)
head(flipsim)
```

Below, we'll plot the results of this simulation for ease of interpretation

```{r}
#create plot
fair_plot <- flipsim %>% ggplot(aes(y = Pct_Heads, x = Trial)) + ggtitle("Percentage of Heads \n Fair Coin") + geom_line() 

#note the below code creates the animation, but it does not look quite correct to me when it animates so I do not use it here. If you want to use this, copy the code after the hashtag (starting with a +) to the line above, just after the geom_line() 

#+ geom_segment(aes(xend = 5000, yend = Pct_Heads), linetype = 2,color = "red") + geom_point(size = 2) + transition_reveal(Trial) + ylim(0,1) + coord_cartesian(clip = "off") + theme(plot.title = element_text(hjust = 0.5)) 

fair_plot
```

## Premier League Goals and the Poisson Distribution

Note, much of the idea and code for this section comes from: <https://bookdown.org/theqdata/honors_thesis/goal-scoring-and-the-poisson-process.html>

The data comes from: <https://www.football-data.co.uk/englandm.php> and is the EPL results for 21-22 Season

However, I admit that for the sake of time, I created in Excel a new table of Total Goals (TOTG) rather than here in R.

```{r}
EPL<-read_excel("D:/Dropbox/R_Files/Data/EPL21-22.xlsx")
head(EPL)
```

```{r}
summary(EPL$HomeTeam)
```

The above is a quick way of calculating / checking the total number of games if you don't know already.

```{r}
summary(EPL$TOTG)
```

```{r}
sum(EPL$TOTG)
```

First, let's plot the distribution of goals.

If you've ever seen a Poisson distribution, you'll recognise this!

```{r}
ggplot(EPL, aes(x = TOTG)) +
  geom_histogram(color = "darkgreen", fill = "lightgreen", bins = 10) +
  scale_x_continuous(breaks= 0:9)
```

So, this looks fairly Poisson-y, but it's always a good idea to check this in a bit more depth. In fact, there is a formal test of this, but we don't need to go to that depth here.

First, let's create a table of all the matches with different numbers of goals.

```{r}
GoalsTable <- 
  EPL %>% 
  group_by(TOTG) %>% 
  summarise(ActualMatches = n())
GoalsTable 
```

Let's pull some stats of the Total Goals variable.

```{r}
fav_stats(EPL$TOTG)
```

Below is some code which will help us to create the distributions and also check them.

```{r}
MeanGoals <- fav_stats(EPL$TOTG)[[6]]
numMatches <- fav_stats(EPL$TOTG)[[8]]
StDevGoals <- fav_stats(EPL$TOTG)[[7]]
VarianceGoals <- StDevGoals ^ 2
```

Importantly, because the Poisson distribution is described only by its mean, the first and most basic check we can do is whether the mean and the variance of the variable are the same (or at least very close)

```{r}
MeanGoals
```

```{r}
VarianceGoals
```

This is not exactly the same, but good enough to be going on with.

So, we are happy that we can try to use the Poisson distribution to describe the number of total goals scored in a premier league game.

The first thing I am going to do is build a figure which compares the actual numbers of goals scored in games with the predicted numbers which would be scored if the goals scored were a perfect Poisson distribution.

This is really just for illustration purposes, not to actually answer our core question, which we will come to...

Ideally, we would build the Poisson probabilities for 0-9 goals (which is our maximum number scored in 21-22). However, if you do this, you will find that the two tables (actual and predicted goals) will have different numbers of rows. This is because there were no games with 8 goals scored in 21-22, but one with 9.

This is a bit annoying, but not a big issue. What we do is build a Poisson probability distribution for 0-8 goals, and treat the final probability as that for '8 or more' goals, for the purposes of drawing our figure. This isn't strictly correct, because we have not collected up our actual goal data into that category, but we could do that if we wanted to.

Anyway, later, we'll break this down properly to answer the question of whether we should predict a game with 10 goals in the 22-23 season, but for our showy graph we don't really need to do it.

```{r}
PoisProb <- dpois(c(0:8), MeanGoals)
POIS <-data.frame(PoisProb)
POIS
```

So, this allows us to make some predictions of what we could expect.

Remembering again that there are 380 games in a season, we can see that there is a 0.16% chance of seeing 4 goals in a game, which equates to 380 x 0.16 = 60.8 games we would expect to see in a season with 4 goals.

We can check this out by creating a new table comparing actual with predicted values...

```{r}
NewGoalsTable <- cbind(GoalsTable, PoisProb) 
NewGoalsTable <- mutate(GoalsTable, 
                        ExpectedMatches = round(numMatches * PoisProb))
NewGoalsTable
```

Remember, as you can see above, the row for '8' is missing, and it goes straight to 9. So, to repeat, just take (for simplicity's sake here) the final row of Expected Matches as meaning '2 games with 8 or more goals'

Anyway, it's pretty scarily close! E.g. in 21-22 there were 60 matches with 4 goals, and that is identical to the predicted amount (the difference with the predicted 60.8 from the probability distribution is likely a rounding error).

Let's plot this for the payoff:

```{r}
NewGoalsTable %>% 
  gather(ActualMatches, ExpectedMatches, 
         key = "Type", value = "numMatches") %>% 
  ggplot(aes(x = TOTG, y = numMatches, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge")
```

So, the above was a bit of show, but what we want is the 'go'. In other words, should we place a bet that there will be a game in the 22-23 season with 10 goals?

Remember, there were 380 games in 21-22, none of which contained 10 goals. So, for ease of thinking about this, let's make the assumption that the PL started in 2021, and there were no games before to count (in reality, we would go back to the last time 10 goals were scored and count from there, but let's go with this in the first instance, and we'll expand later on, promise).

So, first, we need to calculate a new set of probabilities out to 10 goals.

```{r}
PoisProb <- dpois(c(0:11), MeanGoals)
POIS <-data.frame(PoisProb)
POIS
```

So, the number we want is the second-last probability: 0.00052.

This is because the table starts from the probability of 0 goals.

This means we can expect a game with 10 goals to happen every 1923 games

Calculate this by dividing 1 by the probability.

So, given there are 380 games per season, you would expect a game with 10 goals to happen once every 5 seasons.

So, if I started counting from the 21-22 season, the answer is NO, I would not bet on there being a game with 10 goals in the 22-23 Premier League season.

But, let's add some further context, as of November 2022, there have been 5 games in the history of the Premier League where 10 goals have been scored (and 1 with 11).

The Premier League has been going since 1992, and so far in Nov 22 there have been...30 seasons. So, we are probably due one.

AND, the last game with 10 goals was in 2013 (Man Utd 5, West Brom 5)

So, I reckon we are definitely-maybe-probably due one.

I might have a flutter after all...

By the way, as of the start of the 22-23 season, there had been 21 matches with 9 goals. We would expect given our Poisson distribution that a game with 9 goals should happen every 1.4 seasons, meaning over 30 seasons we would expect.... 21.

**2023 Update** 
So, the above example is *totally unchanged* from when I originally coded it in 2022. The million dollar question: *was there a game with 10 goals in the 22-23 season...?*
The answer is no, there wasn't. But there were two with 9! Luckily I am not a betting man. I do think we are due one, so this season is probably a good bet... Unless of course you think something has fundamentally changed about the premier league in recent years, changing the probability of goals (and thus suggesting recent seasons are drawn from a *Poission distribution with different characteristics* than past ones), which is in fact an interesting question...
